---
title: Supplement to The Evolving Roles of Partisanship and Vulnerability in the COVID-19
  Pandemic
output: 
  pdf_document:
    keep_tex: true
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  out.width = '70%',
  out.height = '40%',
  fig.align = 'center')

library(here)
library(knitr)
library(cowplot)
library(patchwork)
```

## Introduction

These supplemental materials present a conditional auto-regressive (CAR) 
Poisson model of COVID-19 county death rates in the United States.  The
inclusion of a conditional auto-regressive spatial component in this model is
intended to address the question of how robust the relationships are between
observed covariates and COVID-19 death rates given the expected correlations in
rates between counties which neighbor one another. 

## Methods 

The implementation of our model is based on the Stan Case Study, *Exact Sparse
CAR Models in Stan*  [@exact-sparse-car]. Spatial models including CAR models
and improvements on the Besag-York-Mollié model have often been used in current
epidemiology and disease risk mapping applications to distinguish spatially
structured effects from the effects of observed covariates [@lee_bayesian_2014;
@wakefield_disease_2007; Morris et al. (2019)].

To describe our model, we write the deaths observed as $y_1, y_2, ..., y_{2683}$
for each of the 2,683 counties which have neighboring counties and for which all
covariates were available and hence were considered in our main manuscript. 
Letting $X_i$ for $i$ in 1...2,683 represent the vector of observed covariates for 
the $i$th county and similarly $P_i$ represent the population of the $i$th county,
we write that 

$$y_i \sim \text{Poisson}(\text{exp}(X_i \beta + \phi_i + \text{log}(P_i))),$$

where $\beta$ is a vector of the estimated coefficients for the covariates and
$\phi_i$ is the the spatial component of the model. See @exact-sparse-car for
the details of the prior distributions on $\beta$ and $\phi$. This model is
fit twice with data from period 2 and period 3 separately.

Given the computational complexity in fitting these models with over 2600 county
observations and a large number of covariates, we opted to only include a
selection of the variables which had the highest measures of feature importance
in the LASSO and spatial linear models from our main manuscript. In order to 
include parameters parsimoniously, we chose to include parameters which 
appeared in the top three most important features from the LASSO and spatial 
linear models for periods 2 and 3. Since we only modeled the probability of 
counties being seeded during period 1 and did not model death rates, we 
have only calibrated the spatial model presented here to the deaths data from
periods 2 and 3.

All analyses in these supplemental materials were conducted using R version
4.0.2 [@R-base] and the model analyses were conducted in the Bayesian
statistical computing and modeling framework Stan using the No-U-Turns
Hamiltonian Monte Carlo Sampler [@Stan;
@homan_no-u-turn_2014].

\newpage

## Results

### Parameter Estimates

We found that the results of fitting a spatial sparse CAR Poisson model were
consistent with our findings from the main analyses for periods 2 and 3.

```{r, out.width='100%', fig.cap="Non-spatial parameter estimates for periods 2 and 3"}
theme_set(
  cowplot::theme_cowplot() +
    theme(text = element_text(family = '')))

p1 <- ggdraw() + 
  draw_image(here('figures/stan_car_poisson_period_2.png'))

p2 <- ggdraw() + 
  draw_image(here('figures/stan_car_poisson_period_3.png'))

p <- plot_grid(p1, p2)

print(p)
```

These model estimates reflect the associations between county-level covariates and
COVID-19 death rates after accounting for the estimated spatial correlation structure included in the model.

We found that higher percentages of residents in nursing homes, Black
non-Hispanic population percentages, and Hispanic population percentages at the
county level were associated with higher COVID-19 death rates during period
2. Counties with higher high school graduation rates were associated with having
lower COVID-19 death rates in period 2. During period 3, counties where the
population voted more Republican (positive political lean), and counties with
higher percentages of the population living in nursing homes were associated
with higher COVID-19 death rates. Counties with higher median income, greater
high school graduation rates, and policy strictness during period 3 were found
to be have lower COVID-19 death rates.

\newpage

### Spatial Components

We present the spatial model component below visualized as 
$\text{exp}(\phi_i + \text{log}(P_i))$, the expected COVID-19 deaths per capita
in each period conditioning out the effects from covariates.

```{r, out.width='80%', out.height = '60%', fig.cap='Spatial model component during periods 2 and 3'}
p1 <- ggdraw() + 
  draw_image(here('figures/spatial_terms_p2.png'))
p2 <- ggdraw() + 
  draw_image(here('figures/spatial_terms_p3.png'))
p <- plot_grid(p1, p2, nrow=2)
print(p)
```

Supplemental Figure 2 allows us to visualize the estimated spatial correlation
structure and how neighboring counties tend to be correlated with one another. 
In essence, we expect to see that counties which have high rates are surrounded 
by counties that also have high rates and vice-versa for low rate counties.

\newpage

### Model Convergence Diagnostics

```{r, out.width = '60%', out.height='25%', fig.cap = 'Traceplots for non-spatial parameters'}
p1 <- ggdraw() + 
  draw_image(here('figures/traceplot_period_2.png'))

p2 <- ggdraw() + 
  draw_image(here('figures/traceplot_period_3.png'))

p <- plot_grid(p1, p2)
print(p)
```

Contemporary advice recommends that Bayesian models should be considered 
to have converged only if the Markov chains have convergence diagnostics of
$\hat{R} < 1.05$ [@rhat; @vehtari_rank-normalization_2020]. In Supplemental
Figure 4 we present the $\hat{R}$ convergence diagnostics for
our non-spatial and spatial effects for both periods 2 and 3 which are all 
below 1.05. 

```{r, out.height='40%', out.width = '100%', fig.cap = 'Convergence diagnostics for all parameters'}

p1 <- ggdraw() + 
  draw_image(here('figures/stan_car_poisson_period_2_rhats.png'))

p2 <- ggdraw() + 
  draw_image(here('figures/stan_car_poisson_period_3_rhats.png'))


p3 <- ggdraw() + 
  draw_image(here('figures/stan_car_poisson_period_2_spatial_rhats.png'))

p4 <- ggdraw() + 
  draw_image(here('figures/stan_car_poisson_period_3_spatial_rhats.png'))

p <- plot_grid(p1, p2, p3, p4, nrow=2, rel_heights = c(0.5, 1.5))
print(p)
```

\newpage

## References